Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-15 paper.tex
about to edit numerical

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

%\title{Linear analysis of the vertical shear instability in astrophysical disks}
\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include a thermal
  relaxation term that restores the disk  
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI at tens of AU with 
  a characteristic growth time of $\sim 30$ local orbital
  periods, and the dominant disturbance in the MMSN is the `corrugation' mode
  which casues the disk column to oscillate vertically. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This dependence
occurs when the equilibrium pressure depends on both the temperature
and density, making the disk baroclinic. It is thus generally
expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple interpretation for the VSI: an
exchange of fluid parcels while conserving their angular momenta
individually, may result in a new configuration of lower energy, if
the exchange occurs over a sufficiently large vertical displacement.

%history

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}. Numerical simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13} and \cite{stoll14}. In these simulations the authors
indeed observe the growth of vertically-global but radially-localized
perturbations when the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
disk temperature essentially needs to remain unchanged during its
evolution. This was further supported with radiation hydrodynamic
simulations by \cite{stoll14}, who find the VSI is only sustained with
the inclusion of heating from stellar irradiation, which resulted in
an effective thermal timescale $t_c\ll P_\mathrm{orb}$.

%why linear approach? computationally fast (need very high resolution) 
%can be used as a first estimate

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually
assumed for typical disk models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk become turbulent through the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of earlier 
analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
evolution, but considered vertically localized disturbances. Our 
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescales.

%paper organization

%say "grossly simplified

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially-local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic thermal timescale below which
the VSI is expected to operate. In \S\ref{numerical} we numerically
solve the linearized fluid equations in the radially-local
approximation, but relax all other assumptions made in our analytical approach.   
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebulae (MMSN) by modeling thermal relaxation
timescales based on realistic disk properties, and find that the VSI may
operate in the outer parts of the MMSN. We summarize and discuss our
results in \S\ref{summary}, including caveats and possible directions
for future work.

@@

%\acknowledgements
\appendix
\input{appendix}

>>>>

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from 
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which casues the disk column to oscillate
  vertically. We suggest the VSI is dynamically imporant in the outer
  parts of protoplanetary disks. 
\end{abstract}

\section{Introduction}
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. This
dependence occurs when the equilibrium pressure depends on both the
temperature and density, making the disk baroclinic. It is therefore
generally expected that $\p_z\Omega\neq 0$ in astrophysical disks.

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial 
lengthscales) to become unstable. \cite{umurhan13} gives a simple
interpretation for the VSI: an exchange of fluid parcels while
conserving their angular momenta individually, may result in a new
configuration of lower energy, if the exchange occurs over a
sufficiently large vertical displacement. In practice, the VSI
corresponds to unstable inertial waves.  
%inertial waves 
%history

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}.  Numerical simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13} and \cite{stoll14}. In these new simulations the authors
indeed observe the growth of vertically global but radially localized
perturbations when the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions, required for the VSI, are stabilized by
buoyancy if the disk is sub-adiabatically stratified for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
disk temperature essentially needs to remain unchanged during its
evolution. This was further supported with radiation hydrodynamic
simulations by \cite{stoll14}, who find the VSI is only sustained with
the inclusion of heating from stellar irradiation, which resulted in
an effective thermal timescale $t_c\ll P_\mathrm{orb}$.

From these studies it is clear that the VSI is sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk models,  
it is important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the relevance of
the VSI in realistic disks. Should a given disk become turbulent
through the VSI, it must first pass through the
linear instability. Thus, understanding the thermodynamic dependence
of the linear VSI is essential.

In this work we study the linear VSI in radially local,
vertically global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations and therefore did not include an energy
equation. Our models are also an extension of earlier 
analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
evolution, but considered vertically localized disturbances. Our 
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescales.

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic upper limit to the thermal
timescale for the VSI to operate. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our linear models to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal relaxation
timescales based on realistic disk properties, and find that the VSI may
operate in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.

@@

\appendix
\input{appendix}
